---
output: html_document
---
# Statistical Inference 
### Course Project - Assignment 2
### Jim Leach

********************************************************************************

### Synopsis

This report has been created for the the *Statistical Inference* MOOC from Johns Hopkins university on [Coursera](https://class.coursera.org/statinference-005).

The second assignment required analysis of the ToothGrowth data set in R. Specifically, the following points had to be addressed:

1. Load the ToothGrowth data and perform some basic exploratory data analyses 
2. Provide a basic summary of the data.
3. Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose.
4. State your conclusions and the assumptions needed for your conclusions. 

See Appendix 1 for the complete, end-to-end R code that generates all the required output.

********************************************************************************

### Data Processing:
```{r packages, echo = FALSE,results = 'hide',message = FALSE}
# i. ggplot2
library(ggplot2,quietly=T)

# ii. magrittr
library(magrittr,quietly=T)

# iii. dplyr
library(dplyr,quietly=T)
```

#### Loading and Summarising ToothGrowth Data
```{r set_parameters, echo = FALSE, results = 'hide', message = FALSE}
# 1i. load the dataset
data(ToothGrowth)

# 1ii. read in to R
data <- ToothGrowth
```

The ToothGrowth data set in R is a small data set detailing the length of odontoblats (teeth) in each of 10 guinea pigs given a dose of Vitamin C in one of three doses (0.5, 1 and 2 mg) and via one of two delivery methods (orange juice or pure ascorbic acid)

The structure of this data is:
```{r str_data, echo=FALSE}
str(data)
```
Looking at this output, this analysis has coded the values of the `supp` variable as follows:
 
* OJ - Orange Juice
* VC - Ascorbic Acid

##### Summarising Data
Given that the `supp` and `dose` variables are values used by the study and not recorded data, it is sufficient to look only at the summary of the `len` data (i.e the tooth length):
```{r summary, echo=FALSE}
summary(data$len)
sd_len <- sd(data$len)
lower <- mean(data$len) - qt(0.975, 59) * sd_len/sqrt(60)
upper <- mean(data$len) + qt(0.975, 59) * sd_len/sqrt(60)
```

Additionally, it is calculated that the standard deviation of the tooth length is `r I(sd_len)`. The 95% confidence interval of the mean ranges from `r round(lower,1)` to `r round(upper,1)`.

#### Exploring the Data
##### Summarising By supplement
Next, a brief summary is performed by supplement:
```{r supp_summary, echo=FALSE}
data_sum_sup <- data %>%
                group_by(supp) %>%
                summarise(mean_growth = mean(len),
                          sd_growth = sd(len),
                          lower_95 = mean_growth - qt(0.975, 29) * 
                                                      (sd_growth/sqrt(30)),
                          upper_95 = mean_growth + qt(0.975, 29) * 
                                                      (sd_growth/sqrt(30)),
                          ci = upper_95 - mean_growth)
                          
data_sum_sup$supp <- ifelse(data_sum_sup$supp == "OJ", "Orange Juice",
                                                       "Ascorbic Acid")
data_sum_sup[,1:5]
```
 
The data presented show the mean growth with orange juice (`r round(data_sum_sup[1,2],2)`) and with pure ascorbic acid (`r round(data_sum_sup[2,2],2)`) as well as the standard deviation and the 95% confidence intervals for these values.

##### Figure 1 - Mean tooth length by supplement:
```{r fig1_mean_length_supp, echo=FALSE}
data_sum_sup %>%
ggplot(aes(x = supp, y = mean_growth), colour = supp)+
  geom_bar(stat = "identity", aes(fill = supp))+
  geom_errorbar(aes(ymin=mean_growth-ci, ymax=mean_growth+ci),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))+
  scale_fill_manual(values = c("darkorange4", "darkorange"))+
  theme_minimal()+
  theme(legend.position = "none")+
  ylab("Mean tooth length")+
  xlab("")
```

It appears that orange juice results in a larger tooth length than pure ascorbic acid. This is tested below.

##### Summarising By Dose
A brief summary is performed by dose:
```{r dose_summary, echo=FALSE}
data_sum_dose <- data %>%
  group_by(dose) %>%
  summarise(mean_growth = mean(len),
            sd_growth = sd(len),
            lower_95 = mean_growth - qt(0.975, 19) * 
              (sd_growth/sqrt(30)),
            upper_95 = mean_growth + qt(0.975, 19) * 
              (sd_growth/sqrt(30)),
            ci = upper_95 - mean_growth)

data_sum_dose[,1:5]
```
The data presented show the mean growth for doses of 2.0mg (`r round(data_sum_dose[1,2],2)`), 1.0mg (`r round(data_sum_dose[2,2],2)`) and 0.5mg (`r round(data_sum_dose[3,2],2)`) as well as the standard deviation and the 95% confidence intervals for these values.

##### Figure 2 - Mean tooth length by dose:
```{r fig2_mean_len_dose, echo=FALSE}
data_sum_dose %>%
ggplot(aes(x = as.character(dose), y = mean_growth), colour = dose)+
  geom_bar(stat = "identity", aes(fill = as.character(dose)))+
  geom_errorbar(aes(ymin=mean_growth-ci, ymax=mean_growth+ci),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))+
  scale_fill_manual(values = c("gray34","gray22","gray10"))+
  xlab("Dosage (mg)")+
  ylab("Mean tooth length")+
  theme_minimal()+
  theme(legend.position = "none")
```

It appears that a higher dose results in a larger tooth length than pure ascorbic acid. This is tested below.

##### Summarising By supplement *and* Dose
A brief summary is performed by supplement *and* dose:
```{r both_summary, echo=FALSE}
data_sum_both <- data %>%
  group_by(supp, dose) %>%
  summarise(mean_growth = mean(len),
            sd_growth = sd(len),
            lower_95 = mean_growth - qt(0.975, 9) * 
              (sd_growth/sqrt(30)),
            upper_95 = mean_growth + qt(0.975, 9) * 
              (sd_growth/sqrt(30)),
            ci = upper_95 - mean_growth)

data_sum_both$supp <- ifelse(data_sum_both$supp == "OJ", "Orange Juice",
                                                         "Ascorbic Acid")

data_sum_both[,1:5]
```
The data presented show the mean growth for doses of 2.0mg, 1.0mg and 0.5mg of the respective supplements as well as the standard deviation and the 95% confidence intervals for these values.

##### Figure 3 - Mean tooth length by supplement and dose:
```{r fig3_mean_len_both, echo=FALSE}
data_sum_both %>%
  ggplot(aes(x = as.character(dose), y = mean_growth), colour = dose)+
  geom_bar(stat = "identity", aes(fill = as.character(dose)))+
  geom_errorbar(aes(ymin=mean_growth-ci, ymax=mean_growth+ci),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))+
  facet_grid(.~supp)+
  scale_fill_manual(values = c("gray34","gray22","gray10"))+
  xlab("Dosage (mg)")+
  ylab("Mean tooth length")+
  theme_minimal()+
  theme(legend.position = "none")
```

It is interesting to note that whilst orange juice appears to result in a larger mean growth at low and medium does (0.5 and 1.0mg), at a higher dosage of 2.0mg it appears that the two supplements are roughly equivalent. 

This is tested below.

*******************************************************************************

### Hypothesis Testing:

Broadly speaking, the exploration of the data has presented two main hypotheses:

1. Orange juice supplementation produces a higher mean tooth growth than ascorbic acid; and
2. A higher dosage of either supplement produces a higher mean tooth growth.

#### Testing Based on supplement
The Student's $t$-test output for the comparison of tooth growth with orange juice compared to ascorbic acid is:
```{r oj_vc_test, echo=FALSE}
oj <- data %>% filter(supp == "OJ") %>% select(len)
vc <- data %>% filter(supp == "VC") %>% select(len)

result <- t.test(oj, vc, alternative = "greater", paired = FALSE, var.equal = TRUE)
result
```

With a $p$-value of `r round(result$p.value,3)` and a $t$-confidence interval of `r result$conf[1]` to `r result$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, and so the data show that orange juice produces a higher mean tooth growth than ascorbic acid.

#### Testing Based on Dose
##### Comparing 2.0mg with 0.5mg
The Student's $t$-test output for the comparison of tooth growth with a dose of 2.0mg compared to 0.5mg is:
```{r dose_test, echo=FALSE}
# lets test that with a t.test:
# hypothesis - higher doses provides a higher growth
dose2 <- data %>% filter(dose == 2.0) %>% select(len)
dose1 <- data %>% filter(dose == 1.0) %>% select(len)
dose0.5 <- data %>% filter(dose == 0.5) %>% select(len)

# perform test for 2 being greater than 0.5
two_half <- t.test(dose2, dose0.5, alternative = "greater", paired = F, var.equal = T)

# perform test for 2 being greater than 1
two_one <- t.test(dose2, dose1, atlernative = "greater", paired = F, var.equal = T)

# perform test for 1 being greater than 0.5
one_half <- t.test(dose1, dose0.5, alternative = "greater", paired = F, var.equal = T)

two_half
```
With a $p$-value of `r I(two_half$p.value)` and a $t$-confidence interval of `r two_half$conf[1]` to `r two_half$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, and so the data show that 2.0mg of supplement produces a higher mean tooth growth than 0.5mg of supplement.

##### Comparing 2.0mg with 1.0mg
Similarly, the Student's $t$-test output for the comparison of tooth growth with a dose of 2.0mg compared to 1.0mg is:
```{r two_one_test, echo=FALSE}
two_one
```
With a $p$-value of `r I(two_one$p.value)` and a $t$-confidence interval of `r two_one$conf[1]` to `r two_one$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, and so the data show that 2.0mg of supplement also produces a higher mean tooth growth than 1.0mg of supplement.

##### Comparing 1.0mg with 0.5mg
Finally, the Student's $t$-test output for the comparison of tooth growth with a dose of 1.0mg compared to 0.5mg is:
```{r one_half_test, echo=FALSE}
one_half
```
With a $p$-value of `r I(one_half$p.value)` and a $t$-confidence interval of `r one_half$conf[1]` to `r one_half$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, and so the data show that 2.0mg of supplement also produces a higher mean tooth growth than 1.0mg of supplement.


#### Testing Based on supplement *and* Dose
The previous results show that orange juice produces a higher mean growth than ascorbic acid; and similarly a higher dosage produces a higher mean growth. 

Whether these results can be cross-validated, i.e. does orange juice provide a higher growth rate at *all* doses (recall from the exploratory analysis that this only seemed to be the case at low and mediaum doses) is shown below:

##### Comparing Orange Juice to Ascorbic Acid at 0.5mg
The Student's $t$-test output is:
```{r oj_vc_half, echo=FALSE}
doseoj.5 <- data %>% filter(supp == "OJ" & dose == .5) %>% select(len)
dosevc.5 <- data %>% filter(supp == "VC" & dose == .5) %>% select(len)
oj_vc_half <- t.test(doseoj.5, dosevc.5, paired = F, var.equal = T)
oj_vc_half
```
With a $p$-value of `r I(oj_vc_half$p.value)` and a $t$-confidence interval of `r oj_vc_half$conf[1]` to `r oj_vc_half$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, i.e. at low doses there appears to be a statistically significant difference between the two supplements.

##### Comparing Orange Juice to Ascorbic Acid at 1.0mg
The Student's $t$-test output is:
```{r oj_vc_1, echo=FALSE}
doseoj1 <- data %>% filter(supp == "OJ" & dose == 1.0) %>% select(len)
dosevc1 <- data %>% filter(supp == "VC" & dose == 1.0) %>% select(len)
oj_vc_1 <- t.test(doseoj1, dosevc1, paired = F, var.equal = T)
oj_vc_1
```
With a $p$-value of `r I(oj_vc_1$p.value)` and a $t$-confidence interval of `r oj_vc_1$conf[1]` to `r oj_vc_1$conf[2]` (i.e. not containing 0) it seems reasonable to reject the null hypothesis, i.e. at medium doses there appears to be a statistically significant difference between the two supplements.


##### Comparing Orange Juice to Ascorbic Acid at 2.0mg
The Student's $t$-test output is:
```{r oj_vc_2, echo=FALSE}
doseoj2 <- data %>% filter(supp == "OJ" & dose == 2.0) %>% select(len)
dosevc2 <- data %>% filter(supp == "VC" & dose == 2.0) %>% select(len)
oj_vc_2 <- t.test(doseoj2, dosevc2, paired = F, var.equal = T)
oj_vc_2
```
With a $p$-value of `r I(oj_vc_2$p.value)` and a $t$-confidence interval of `r oj_vc_2$conf[1]` to `r oj_vc_2$conf[2]` (i.e. *containing* 0) it seems reasonable to *fail* to reject the null hypothesis, i.e. at higher doses there does not appear to be a statistically significant difference between the two supplements.

*******************************************************************************
### Assumptions

All the Student's $t$-tests in this exercise have been conducted assuming equal variances. As all subjects were guinea pigs, it was assumed that the population was simply *all guinea pigs* and so the *population* variance in tooth size would be equal in each subject group.

Additionally, in order to perform a Students $t$-test, it is assumed that the data are normally distributed. With a sample of only `r nrow(data)` guinea pigs, this is a reasonable assumption to make. However it is imporant to **test this assumption**. This is done by plotting a density distribution of the sample, and overlaying a normal distribution curve:

```{r assumption_test, echo=FALSE}
data %>%
  ggplot(aes(x = len))+
  geom_density(fill = "gray10", alpha = 0.3, color= "white")+
  theme_minimal()+
  xlab("Tooth Length")+
  ylab("Density")+
  ggtitle("Density Plot of Tooth Size")+
  theme(panel.grid.minor.y=element_blank(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank())+
  stat_function(color="darkorange",
                fun= dnorm, 
                args = list(mean = mean(data$len), sd = sd(data$len)))
```

Whilst this plot shows that it is the case that the data deviate *slightly* from the truly normal distribution with a small negative skew, it seems reasonable to maintain this assumption as the deviation is very small.

*******************************************************************************
### Conclusions

Overall, it is possible to conclude that there is a statistically significant difference between both orange juice and ascorbic acid, and the dosage at which these supplements are applied.

Generally speaking a higher dosage results in a larger tooth growth and similarly orange juice results in a larger growth than pure ascorbic acid. 

However, whilst this is true intra-supplement and medium and low does, at high doses, the two suppliments appear to produce approximately the same growth, with no statistically significant difference between the two.

*******************************************************************************
### References and Contact

A number of packages were used and are therefore required for this analysis. These packages were:

* [magrittr](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)
* [ggplot2](http://ggplot2.org/)
* [dplyr](http://blog.rstudio.org/2014/01/17/introducing-dplyr/)

The complete code and documentation for this assignment can be found on [GitHub](https://github.com/Jim89/stat_infer) repository. 

The author of this report can be contacted on [twitter](https://twitter.com/leach_jim) 

********************************************************************************

### Appendix 1 - Complete R Code

    ################################################################################
    # Assignment 2 - Analysis of ToothGrowth data set
    ################################################################################
    # Load ToothGrowth data and perform basic exploratory analysis
    # basic summary of data
    # confidence intervals and hypothesis tests to compare growth by supp and dose
    # state conclusions and assumptions
    
    ################################################################################
    # Step 0  - load packages
    ################################################################################
    library(ggplot2,quietly=T)
    library(magrittr,quietly=T)
    library(dplyr,quietly=T)
    
    ################################################################################
    # Step 1  - load data and perform some brief summaries
    ################################################################################
    data(ToothGrowth)
    data <- ToothGrowth
    
    summary(data$len)
    sd_len <- sd(data$len)
    
    ################################################################################
    # Step 2  - explore based on suppliment
    ################################################################################
    # descriptive stats per suppliment
    data_sum_sup <- data %>%
      group_by(supp) %>%
      summarise(mean_growth = mean(len),
                sd_growth = sd(len),
                lower_95 = mean_growth - qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                upper_95 = mean_growth + qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                ci = upper_95 - mean_growth)
    
    data_sum_sup$supp <- ifelse(data_sum_sup$supp == "OJ", "Orange Juice",
                                "Ascorbic Acid")
    
    data_sum_sup %>%
    ggplot(aes(x = supp, y = mean_growth), colour = supp)+
      geom_bar(stat = "identity", aes(fill = supp))+
      geom_errorbar(aes(ymin=mean_growth-ci, ymax=mean_growth+ci),
                    width=.2,                    # Width of the error bars
                    position=position_dodge(.9))+
      scale_fill_manual(values = c("darkorange4", "darkorange"))+
      theme_minimal()+
      theme(legend.position = "none")+
      ylab("Mean tooth length")+
      ggtitle("Mean tooth length by suppliment administered")+
      xlab("")
    
    ################################################################################
    # Step 3  - explore based on dose
    ################################################################################
    # average length per dose per suppliment
    data_sum_dose <- data %>%
      group_by(dose) %>%
      summarise(mean_growth = mean(len),
                sd_growth = sd(len),
                lower_95 = mean_growth - qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                upper_95 = mean_growth + qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                ci = upper_95 - mean_growth)
    
    data_sum_dose %>%
      ggplot(aes(x = as.character(dose), y = mean_growth), colour = dose)+
      geom_bar(stat = "identity", aes(fill = as.character(dose)))+
      geom_errorbar(aes(ymin=mean_growth-ci, ymax=mean_growth+ci),
                    width=.2,                    # Width of the error bars
                    position=position_dodge(.9))+
      scale_fill_manual(values = c("gray34","gray22","gray10"))+
      xlab("Dosage (mg)")+
      ylab("Mean tooth length")+
      theme_minimal()+
      theme(legend.position = "none")
    
    ################################################################################
    # Step 4  - explore based on suppliment *and* dose
    ################################################################################
    # average length per dose per suppliment
    data_sum_both <- data %>%
      group_by(supp, dose) %>%
      summarise(mean_growth = mean(len),
                sd_growth = sd(len),
                lower_95 = mean_growth - qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                upper_95 = mean_growth + qt(0.975, 29) * 
                  (sd_growth/sqrt(30)),
                ci = upper_95 - mean_growth)
    
    data_sum_both$supp <- ifelse(data_sum_both$supp == "OJ", "Orange Juice",
                                 "Ascorbic Acid")
    
    data_sum_both %>%
      ggplot(aes(x = as.character(dose), y = mean_growth), colour = dose)+
      geom_bar(stat = "identity", aes(fill = as.character(dose)))+
      facet_grid(.~supp)+
      scale_fill_manual(values = c("gray34","gray22","gray10"))+
      xlab("Dosage (mg)")+
      ylab("Mean tooth length")+
      ggtitle("Mean tooth length by dose (mg) administered")+
      theme_minimal()+
      theme(legend.position = "none")
    
    ################################################################################
    # Step 5  - hypothesis testing based on suppliment
    ################################################################################
    # lets test that with a t.test:
    # hypothesis - orange juice provides a higher growth
    oj <- data %>% filter(supp == "OJ") %>% select(len)
    vc <- data %>% filter(supp == "VC") %>% select(len)
    
    t.test(oj, vc, alternative = "greater", paired = FALSE, var.equal = TRUE)
    
    ################################################################################
    # Step 6  - hypothesis testing based on dose
    ################################################################################
    # lets test that with a t.test:
    # hypothesis - higher doses provides a higher growth
    dose2 <- data %>% filter(dose == 2.0) %>% select(len)
    dose1 <- data %>% filter(dose == 1.0) %>% select(len)
    dose0.5 <- data %>% filter(dose == 0.5) %>% select(len)
    
    # perform test for 2 being greater than 0.5
    t.test(dose2, dose0.5, alternative = "greater", paired = F, var.equal = T)
    
    # perform test for 2 being greater than 1
    t.test(dose2, dose1, atlernative = "greater", paired = F, var.equal = T)
    
    # perform test for 1 being greater than 0.5
    t.test(dose1, dose0.5, alternative = "greater", paired = F, var.equal = T)
    
    
    ################################################################################
    # Step 7  - hypothesis testing based on suppliment and dose
    ################################################################################
    doseoj2 <- data %>% filter(supp == "OJ" & dose == 2.0) %>% select(len)
    dosevc2 <- data %>% filter(supp == "VC" & dose == 2.0) %>% select(len)
    t.test(doseoj2, dosevc2, paired = F, var.equal = T)
    
    doseoj1 <- data %>% filter(supp == "OJ" & dose == 1.0) %>% select(len)
    dosevc1 <- data %>% filter(supp == "VC" & dose == 1.0) %>% select(len)
    t.test(doseoj1, dosevc1, paired = F, var.equal = T)
    
    doseoj.5 <- data %>% filter(supp == "OJ" & dose == .5) %>% select(len)
    dosevc.5 <- data %>% filter(supp == "VC" & dose == .5) %>% select(len)
    t.test(doseoj.5, dosevc.5, paired = F, var.equal = T)
    
    ################################################################################
    # Step 8  - test assumptions - data normally distributed
    ################################################################################
    data %>%
      ggplot(aes(x = len))+
      geom_density(fill = "gray10", alpha = 0.3, color= "white")+
      theme_minimal()+
      xlab("Tooth Length")+
      ylab("Density")+
      ggtitle("Density Plot of Tooth Size")+
      theme(panel.grid.minor.y=element_blank(),
            panel.grid.major.x=element_blank(),
            panel.grid.minor.x=element_blank())+
      stat_function(color="darkorange",
                    fun= dnorm, 
                    args = list(mean = mean(data$len), sd = sd(data$len)))




